Tail Flexibility in the Degrees of Preferential Attachment Networks

### 2025-05-19 paper.tex
Restructuring paper

% Options for packages loaded elsewhere
% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  sn-basic,
]{sn-jnl}

\usepackage{xcolor}
\usepackage{amsmath,amssymb}
\setcounter{secnumdepth}{5}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
% Make \paragraph and \subparagraph free-standing
\makeatletter
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}{
    \@ifstar
      \xxxParagraphStar
      \xxxParagraphNoStar
  }
  \newcommand{\xxxParagraphStar}[1]{\oldparagraph*{#1}\mbox{}}
  \newcommand{\xxxParagraphNoStar}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}{
    \@ifstar
      \xxxSubParagraphStar
      \xxxSubParagraphNoStar
  }
  \newcommand{\xxxSubParagraphStar}[1]{\oldsubparagraph*{#1}\mbox{}}
  \newcommand{\xxxSubParagraphNoStar}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\makeatother

\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\newsavebox\pandoc@box
\newcommand*\pandocbounded[1]{% scales image to fit in text height/width
  \sbox\pandoc@box{#1}%
  \Gscale@div\@tempa{\textheight}{\dimexpr\ht\pandoc@box+\dp\pandoc@box\relax}%
  \Gscale@div\@tempb{\linewidth}{\wd\pandoc@box}%
  \ifdim\@tempb\p@<\@tempa\p@\let\@tempa\@tempb\fi% select the smaller of both
  \ifdim\@tempa\p@<\p@\scalebox{\@tempa}{\usebox\pandoc@box}%
  \else\usebox{\pandoc@box}%
  \fi%
}
% Set default figure placement to htbp
\def\fps@figure{htbp}
\makeatother

\setlength{\emergencystretch}{3em} % prevent overfull lines

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

@@

\raggedbottom
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\usepackage{amsthm}
\theoremstyle{plain}
\newtheorem{proposition}{Proposition}[section]
\theoremstyle{remark}
\AtBeginDocument{\renewcommand*{\proofname}{Proof}}
\newtheorem*{remark}{Remark}
\newtheorem*{solution}{Solution}
\newtheorem{refremark}{Remark}[section]
\newtheorem{refsolution}{Solution}[section]
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Tail Flexibility in the Degrees of Preferential Attachment Networks},
  pdfauthor={Thomas William Boughen; Clement Lee; Vianey Palacios Ramirez},
  pdfkeywords={networks, extremes},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

The degree distribution of a network can be very informative about the
networks structure, giving some understanding about the inequality in
the influence of the nodes. Often the degree distribution is one of the
simplest places to look when trying to understand this, as the full
evolution of the network is usually not available. This makes modelling
the degree distribution an important topic in research with
contributions such as {[}ref{]}.

When looking at the degree distributions of real networks, a power law
is an attractive option to use when modelling as it is seemingly
observed in a lot of networks. Fitting the discrete power law to the
degree distribution come with the added benefit of suggesting
preferential attachment as the generating mechanism since it has been
shown to generate networks with a power law degree distribution.

If the power law is an adequate model to use for real degree
distributions has become a heated debate, with alternatives such as
\citet{Broido_2019} seeming to outperform the power law in some cases.
Additionally, \citet{lef24} find that for many real degree distributions
the body may follow a power law, the right tail does not or at least
seems to be lighter than is implied by the body whilst still being
heavy. One method for modelling the tail of the degree distributions of
real networks is to use a discretised variation of the generalised
Pareto distribution (GPD) usually dubbed the Integer GPD (GPD) where
\(X|X>v \sim \mathrm{IGPD}(\xi,\sigma,v)\) is defined by the survival:

@@

\citep[CL: Is the precise form of the IGPD needed here? The big picture
is probably more important: extreme value methods have not been used in
analysing degree distributions until recently, such as][ and
\citet{lef24}. Not using these methods essentially downweights the
deviation of the vertices with large degrees, which have far more
influence on the network than those with small degrees, from the power
law behaviour.]{vvvk19}

\citet{shimura12} introduces a quantity that will help in determining
what domain of attraction a discrete distribution belongs to, that is:

\citet{shimura12} then states that if
\(\lim_{n\rightarrow\infty} \Omega(F,n) = 1/\alpha\) (\(\alpha>0\)),
then \(F\) is heavy tailed with \(\bar F(n) \sim n^{-\alpha}\).
Additionally, if \(\lim_{n\rightarrow\infty} \Omega(F,n) = 0\) then the
distribution is light tailed.

\citep[CL: The results of][ could be moved closer to Proposition 2.1. In
the introduction, if you want to include \citet{shimura12}, you can
mention the paper in passing, when outlining your approach, e.g.~``Using
the results by \citet{shimura12}, we derive the tail index / heaviness
of the limiting degree distribution of the GPA model \citep{rudas07}.
This informs us the precise class of preference functions that result in
flexible heavy tail.'']{shimura12}

These models that seem to outperform the power law do not however come
with the added benefit of suggesting a generating mechanism for the
networks in the same way the power law does. Efforts to make the
preferential attachment model more flexible have not yet remedied this
issue with \citet{krapivsky01} finding that using a sub/super-linear
preference function yields a Weibull tail and a degenerate network,
where one vertex eventually gains all new edges, respectively.

The network generative model that we will be focussing on in this paper
is dubbed General Preferential Attachment in \citet{rudas07} and is
defined as follows:

Starting at time \(t=0\) with an initial network of \(m\) vertices that
each have no edges, at times \(t=1,2,\ldots\) a new vertex is added to
the network bringing with it \(m\) directed edges (with the new vertex
as the source); the target for each of these edges are selected from the
vertices already in the network with weights proportional to some
function \(b\) of their degree.

@@

Estimating the preference function used by network generative models
like this is of great interest, with the potential to provide a deeper
understanding of the relationships between the vertices in the network.
Several papers and at least one R package (\texttt{PAfit}) have been
dedicated to estimating the preference function that may have been used
when a network was growing; one thing all these have in common is
requiring access to a networks evolution data, something that is not
always readily available or feasible to access. A method for estimating
a preference function without this data available would be extremely
useful.

\citep{rudas07} introduces theoretical results for this model with a
general preference function in the cases that the network being
generated is a tree (\(m=1\)); presenting an opportunity to design a
preference function that will capture the behaviours observed in the
degree distributions of real networks. Since they provide theoretical
results for the limiting degree distribution, a model of this form would
be able to be fitted to the degree distributions of real networks where
the full evolution is not available, only a snapshot.

Section~\ref{sec-gpa} looks at the theoretical limiting behaviour of a
GPA model when \(m=1\) according to the results from \citep{rudas07},
going on to introduce a flexible class of preference functions that can
guarantee a heavy tailed degree distribution while still being flexible
in the body. Section~\ref{sec-rec} shows that the preference function
parameters can be fairly well recovered from the degree distributions of
networks simulated from the GPA model using various functions of the
class introduced in Section~\ref{sec-gpa}. Having shown that for
simulated data the model parameters are recoverable from the degree
distribution alone, Section~\ref{sec-real} builds upon this and attempts
to fit the model to degree distributions of real networks; modelling not
only the degree distribution itself but also the estimates for the
parameters of the preference function assuming the networks evolved
according to the GPA scheme. Ultimately, Section~\ref{sec-conc}
discusses the main results of the paper, pitfalls of this method and
future work.

\section{General Preferential Attachment}\label{sec-gpa}

The focus of this section is the limiting degree distribution of a
preferential attachment network with preference function \(b(\cdot)\)
where each vertex brings with it \(m\) edges. This preference function
is subject to the following conditions:

\[
b:\mathbb N \mapsto \mathbb R^+\setminus\{0\},
\]

Given these two conditions an expression for the survival of the
limiting degree distribution can be found in the case that \(m=1\);
obtained by considering a branching process that is equivalent to the
growth of the network, as in \citep{rudas07}. Theorem 1 from
\citep{rudas07} states that for the tree \(\Upsilon(t)\) at time \(t\):

@@

\[
f(k) = \frac{\lambda^*}{\lambda^* + b(k)}\prod_{i=0}^{k-1}\frac{b(i)}{\lambda^*+b(i)},\qquad k\in\mathbb N.
\]

which aligns with the result from \citep{rudas07}, where the expression
is derived using the same branching process results.

Using results from \citep{shimura12} the domain of attraction of this
limiting degree distribution can be found.

@@

Proposition~\ref{prp-omega} implies that the only way to obtain a heavy
tailed degree distribution using a non-decreasing preference function is
to have that preference function be asymptotically linear. The next
subsection focuses on a particular class of these preference functions
that provide flexible behaviour but guarantee a heavy tail.

\subsection{Preferential Attachment with flexible heavy
tail}\label{sec-model}

Previously, in \citep{krapivsky01}, a preference function of the form
\(k^\alpha +\varepsilon\) has been considered. However, it has been
noted that super-linear preferential attachment (\(\alpha > 1\)) leads
to a degenerate limiting degree distribution where at some point one
vertex gains the connection from every vertex that joins the network,
additionally sub-linear preferential attachment (\(\alpha <1\)) has been
shown to lead to a light tailed limiting degree distribution. Consider a
preference function of the form:

@@

@@

\includegraphics[width=0.8\linewidth,height=\textheight,keepaspectratio]{paper_files/figure-pdf/fig-polylinsurv-1.pdf}

@@

The survival (Equation~\ref{eq-polysurv}) can be connected to the IGPD
mentioned in Section~\ref{sec-intro} by using Sterling's approximation
of the gamma function to obtain:

@@

\includegraphics[width=0.8\linewidth,height=\textheight,keepaspectratio]{paper_files/figure-pdf/fig-approx_surv-1.pdf}

@@

\includegraphics[width=0.8\linewidth,height=\textheight,keepaspectratio]{paper_files/figure-pdf/fig-polyheat-1.pdf}

@@

\subsubsection{Piecewise Linear Model}\label{piecewise-linear-model}

When \(\alpha = 1\) the preference function simplifies to a piecewise
linear function with \(\beta\) representing the ratio of the gradients
of the two components:

\[
b(k) = \min(k,k_0) + \varepsilon + \mathbb I\{k\ge k_0\}\beta(k-k_0)
\] for some \(\beta, \varepsilon, k_0>0\).

Following Equation~\ref{eq-polysurv} the limiting degree distribution
is:

\begin{equation}\phantomsection\label{eq-linsurv}{
\bar F(k) = \begin{cases}
\frac{\Gamma(\lambda^*+\varepsilon)\Gamma(k+1+\varepsilon)}{\Gamma(\varepsilon)\Gamma(k+1+\lambda^*+\epsilon)},&k<k_0\\
\frac{\Gamma(\lambda^*+\varepsilon)\Gamma(k_0+\varepsilon)}{\Gamma(\varepsilon)\Gamma(k_0+\lambda^*+\varepsilon)}\times \frac{\Gamma\left(\frac{k_0+\varepsilon+\lambda^*}{\beta}\right)}{\Gamma\left(\frac{k_0+\varepsilon}{\beta}\right)}\times\frac{\Gamma\left(k-k_0 + \frac{k_0+\varepsilon}{\beta}+1\right)}{\Gamma\left(k-k_0 + \frac{k_0+\varepsilon+\lambda^*}{\beta}+1\right)},&k\ge k_0
\end{cases}
}\end{equation}

Just like the previous model, the survival can be approximated by:

\[
\bar F(k) \approx \begin{cases}
\frac{\Gamma(\lambda^*+\varepsilon)}{\Gamma(\varepsilon)} (1+\varepsilon)^{\lambda^*}\left(\frac{k}{1+\varepsilon} + 1\right)^{-\lambda^*},&k<k_0\\
\frac{\Gamma(\lambda^*+\varepsilon)\Gamma(k_0+\varepsilon)}{\Gamma(\varepsilon)\Gamma(k_0+\lambda^*+\varepsilon)}\times\left(\frac{\beta(k+1-k_0)}{k_0+\varepsilon} + 1\right)^{-\lambda^*/\beta},&k\ge k_0
\end{cases}
\]

Some examples of the possible degree distributions are shown in the
central row of Figure~\ref{fig-polylinsurv}.

@@

\includegraphics[width=0.8\linewidth,height=\textheight,keepaspectratio]{paper_files/figure-pdf/fig-rec1-1.pdf}

@@

\includegraphics[width=0.8\linewidth,height=\textheight,keepaspectratio]{mcmc_plot.jpg}

@@

\caption{\label{fig-rec2}Posterior estimates of parameters for data
simulated from the proposed model with various combinations of
(\(\alpha\),\(\beta\),\(\varepsilon\)) and \(k_0=20\).}

@@

Turning now to real data, the goal is to fit the model to the degree
distributions of real networks from various sources. {[}describe
networks being used{]}. Alongside fitting this model to the degree
distributions, we then compare the fit to that of an mixture
distribution that was used in \citet{lef24} and is similar to others
that have been used to model degree distributions {[}others{]}.

@@

\includegraphics[width=0.8\linewidth,height=\textheight,keepaspectratio]{paper_files/figure-pdf/fig-real1-1.pdf}

@@

\includegraphics[width=0.8\linewidth,height=\textheight,keepaspectratio]{paper_files/figure-pdf/fig-shapes-1.pdf}

@@

\includegraphics[width=0.8\linewidth,height=\textheight,keepaspectratio]{paper_files/figure-pdf/fig-pa-1.pdf}

@@

\includegraphics[width=0.8\linewidth,height=\textheight,keepaspectratio]{pars_plot.png}

@@

\caption{\label{fig-rec3}Posterior estimates of parameters for real
data.}

@@

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  sn-basic,
]{sn-jnl}

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{5}
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother

@@

\raggedbottom
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\usepackage{amsthm}
\theoremstyle{plain}
\newtheorem{proposition}{Proposition}[section]
\theoremstyle{remark}
\AtBeginDocument{\renewcommand*{\proofname}{Proof}}
\newtheorem*{remark}{Remark}
\newtheorem*{solution}{Solution}
\newtheorem{refremark}{Remark}[section]
\newtheorem{refsolution}{Solution}[section]
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Tail Flexibility in the Degrees of Preferential Attachment Networks},
  pdfauthor={Thomas William Boughen; Clement Lee; Vianey Palacios Ramirez},
  pdfkeywords={networks, extremes},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

Networks appear in many fields such as sociology, politics,
epidemiology, and economics. Statistical methods have been and continue
to be used to study networks that appear in these fields and beyond,
from the use of stochastic block models to detect communities within the
French political blogosphere \citep{Latouche11}, to using Exponential
Random Graph Models (ERGMs) to analyse the global trade network
\citep{Setayesh22}, and the use of mechanistic models in neuroscience to
investigate the patterns of connections in neural systems
\citep{Betzel17}. One such mechanistic model is the general preferential
attachment (PA) model where as a network grows and new vertices join
they connect to those already present in the network with weights
proportional to some preference function \(b\) of their degree. The
well-known Barabási-Albert (BA) model is a special case of this when
\(b\sim k\) {[}BA99{]}.

Mechanistic models, like the PA model, are in general only particularly
useful when simulating networks and not fitting to the real data as
usually only snapshots of real networks are available and not the entire
evolution. As such, one of the most informative statistics when
regarding a network is the degree distribution. When looking at the
degree distributions of real networks it is attractive to many to use
the power law to model the degree distribution, as the BA model
generates degrees that follow a power law, and therefore claim that
since the degree distribution is roughly power law then the underlying
growth mechanism may be preferential attachment such as the BA model.
However, it has been shown that a general preference function will not
always lead to a power law degree distribution \citep{krapivsky01} and
additionally it has been debated for decades whether most real data
actually follows a power law \citep{Voitalov_2019} or
not\citep{Broido_2019}.

Throughout this debate the use of methods from extremes has been absent
until relatively recently with the papers such as \citep{Voitalov_2019}
and {[}Lee{]}. Forgoing the use of extreme value methods essentially
downplays how deviations from the power law in the largest degrees may
affect the network given that they are often much more influential than
the vertices with smaller degrees.

Noting this, many have now started to use these extreme value methods
when analysing networks. \citep{wan19} showed that extreme value
estimation methods prove more robust than existing parametric approaches
especially when the data are corrupted or the model has been
misspecified, demonstrating that extreme value methods can be a valuable
tool when studying networks. \citep{wang2022random} investigates a PA
model with a affine preference function (similar to the BA model) with
the added behaviour of heterogeneous reciprocal edges, they show that
the distribution of in-degrees and out-degrees produced by such a model
is multivariate regularly varying and they also have hidden regular
variation. {[}Lee{]} using methods from discrete extremes modelled the
right tail of network degrees with a discretised variation of the
generalised Pareto distribution (GPD) usually dubbed the Integer GPD
(IGPD) where \(X|X>v \sim \mathrm{IGPD}(\xi,\sigma,v)\) is defined by
the survival:

@@

The study of degree distributions, whether using extremes or not,
usually does not reveal any information about the preference function
even if PA was the underlying generating mechanism of the network. Here
we address this gap in the literature, more specifically investigating
if given the degree distribution of a network assumed to come from the
PA model can we directly infer the model parameters? We build upon the
foundation provided by \citep{rudas07} deriving the limiting degree
distribution in terms of the preference function and suggest a class of
preference functions that result in realistic tails that are heavy but
are lighter than implied by the power law.

\textbf{?@sec-gpa} looks at the theoretical limiting behaviour of a GPA
model when \(m=1\) according to the results from \citep{rudas07}, going
on to introduce a flexible class of preference functions that can
guarantee a heavy tailed degree distribution while still being flexible
in the body. Section~\ref{sec-rec} shows that the preference function
parameters can be fairly well recovered from the degree distributions of
networks simulated from the GPA model using various functions of the
class introduced in \textbf{?@sec-gpa}. Having shown that for simulated
data the model parameters are recoverable from the degree distribution
alone, Section~\ref{sec-real} builds upon this and attempts to fit the
model to degree distributions of real networks; modelling not only the
degree distribution itself but also the estimates for the parameters of
the preference function assuming the networks evolved according to the
GPA scheme. Ultimately, Section~\ref{sec-conc} discusses the main
results of the paper, pitfalls of this method and future work.

\section{PA model}\label{pa-model}

The network generative model that we will be focussing on in this paper
is dubbed General Preferential Attachment in \citep{rudas07} and is
defined as follows:

Starting at time \(t=0\) with an initial network of \(m\) vertices that
each have no edges, at times \(t=1,2,\ldots\) a new vertex is added to
the network bringing with it \(m\) directed edges (with the new vertex
as the source); the target for each of these edges are selected from the
vertices already in the network with weights proportional to some
function \(b\) of their degree where the preference function \(b\) is
chosen such that:

\begin{equation}\phantomsection\label{eq-condb1}{
b:\mathbb N \mapsto \mathbb R^+\setminus\{0\},
}\end{equation}

@@

Given conditions Equation~\ref{eq-condb1} and Equation~\ref{eq-condb2}
an expression for the survival of the limiting degree distribution can
be found in the case that \(m=1\); obtained by considering a branching
process that is equivalent to the growth of the network, as in
\citep{rudas07}. Theorem 1 from \citep{rudas07} states that for the tree
\(\Upsilon(t)\) at time \(t\):

@@

\[
f(k) = \frac{\lambda^*}{\lambda^* + b(k)}\prod_{i=0}^{k-1}\frac{b(i)}{\lambda^*+b(i)},\qquad k\in\mathbb N.
\] We are specifically interested in how the tail heaviness (\(\xi\) in
the IGPD) is affected by the preference function \(b\),
\citep{shimura12} introduces a quantity that will help in determining
what domain of attraction this discrete distribution belongs to, that
is:

\citep{shimura12} then states that if
\(\lim_{n\rightarrow\infty} \Omega(F,n) = 1/\alpha\) (\(\alpha>0\)),
then \(F\) is heavy tailed with \(\bar F(n) \sim n^{-\alpha}\).
Additionally, if \(\lim_{n\rightarrow\infty} \Omega(F,n) = 0\) then the
distribution is light tailed. This allows us to show the following

@@

See \hyperref[appendix]{appendix} for the details of the proof.

Proposition~\ref{prp-omega} aligns with the result from
\citep{krapivsky01} demonstrating that a sub-linear preference function
will lead to a light tailed distribution. So, in order for the degree
distribution to be heavy tailed we need that the limit
\(\lim_{k\rightarrow\infty} b(k+1)-b(k)\) exists and is positive. We can
in fact show that BA model produces a heavy tailed degree distribution
with index \(\xi=0.5\) by considering the preference function
\(b(k) = k + \varepsilon\), noting that the limit
\(\lim_{k\rightarrow\infty} b(k+1)-b(k)\) is simply
\(\lim_{k\rightarrow\infty} 1 = 1\) leaving the tail heaviness to be
\(1/\lambda^*\) which using \(\hat\rho\) can be found to be \(1/2\).

We can show that in order for the degree distribution to be heavy
tailed, the preference function must be asymptotically linear
i.e.~\(\lim_{k\rightarrow\infty}\frac{b(k)}{k} =  c>0\).

Consider the definition of heavy tails:

\begin{align*}
\lim_{k\rightarrow\infty}\Omega(F,n) &= c>0\\
\lim_{k\rightarrow\infty}\frac{b(k+1)-b(k)}{\lambda^*} &= c\\
\lim_{k\rightarrow\infty}[b(k+1)-b(k)] &= \lambda^*c\\
\lim_{k\rightarrow\infty}\frac{b(k+1)-b(k)}{(k+1)-(k)} &= \lambda^*c\\
\end{align*}

by Stolz-Cesaro Theorem:

\[
\lim_{k\rightarrow\infty}\frac{b(k)}{k} = \lambda^*c > 0.
\]

Since \(b\) must be asymptotically linear in order to have positive tail
heaviness we propose the following preference function:

@@

This preference function, as per Proposition~\ref{prp-omega}, will
produce a degree distribution with tail heaviness \(\beta/\lambda^*\)
guaranteeing a heavy tail. We study this preference function further in
the next section.

\section{Preferential Attachment with flexible heavy
tail}\label{sec-model}

@@

\includegraphics[width=0.8\textwidth,height=\textheight]{paper_files/figure-pdf/fig-polylinsurv-1.pdf}

@@

The survival (Equation~\ref{eq-polysurv}) can be connected to the IGPD
mentioned in Section~\ref{sec-intro} by using Stirling's approximation
of the gamma function to obtain:

\[
\frac{\Gamma(x+y)}{\Gamma(x)}\approx x^{y},
\] and utilising the expression from Equation~\ref{eq-polysurv}:

\begin{align*}
\bar F(k|k\ge k_0) &= \frac{\Gamma\left(\frac{\lambda^* + k_0^\alpha + \varepsilon}{\beta}\right)}{\Gamma\left(\frac{k_0^\alpha + \varepsilon}{\beta}\right)}\times\frac{\Gamma\left(k-k_0  +1 + \frac{k_0^\alpha + \varepsilon}{\beta}\right)}{\Gamma\left(k-k_0  +1 + \frac{\lambda^*+ k_0^\alpha + \varepsilon}{\beta}\right)}\\
&\approx\left(\frac{k_0^\alpha+\varepsilon}{\beta}\right)^{\lambda^*/\beta}\left(k-k_0+1+\frac{k_0^\alpha + \varepsilon}{\beta}\right)^{-\lambda^*/\beta}\\
&=\left(\frac{k_0^\alpha+\varepsilon}{k_0^\alpha+\varepsilon + \beta}\right)^{\lambda^*/\beta}\left(\frac{\beta(k-k_0)}{\beta + k_0^\alpha+\varepsilon} + 1\right)^{-\lambda^*/\beta}\\
&=\left(\frac{\beta(k+1-k_0)}{k_0^{\alpha}+\varepsilon} + 1\right)^{-\lambda^{*}/\beta}
\end{align*}

@@

\includegraphics[width=0.8\textwidth,height=\textheight]{paper_files/figure-pdf/fig-approx_surv-1.pdf}

@@

\includegraphics[width=0.8\textwidth,height=\textheight]{paper_files/figure-pdf/fig-polyheat-1.pdf}

@@

For a given network with degree count vector
\(\pmb n = (n_0, n_1, \ldots, n_M)\) and maximum degree \(M\) the
likelihood is given by:

\begin{align*}
L(\pmb x,\pmb n | \pmb \theta) = \left(\frac{\lambda^*}{\lambda^*+\varepsilon}\right)^{n_0}\left(\prod_{j=l}^{k_0-1}\frac{j^\alpha +\varepsilon}{\lambda^* + j^\alpha +\varepsilon}\right)^{\left(\sum_{i\ge k_0}n_{i}\right)} \prod_{l \le i<k_0}\left(\frac{\lambda^*}{\lambda^* +i^\alpha + \varepsilon } \prod_{j=l}^{k_0-1}\frac{j^\alpha + \varepsilon}{\lambda^* + j^\alpha + \varepsilon}\right)^{n_i}\\ \times \prod_{i\ge k_0}\left(\frac{\text{B}(i-k_0 + (k_0^\alpha + \varepsilon)/\beta,1+\lambda^*/\beta)}{\text{B}((k_0^\alpha + \varepsilon)/\beta,\lambda^*/\beta)}\right)^{n_i}
\end{align*}

where \(B(y,z)\) is the the beta function and \(l\ge0\) variable that
allows truncating the data such that the minimum degree is \(l\).

This likelihood allows for the parameters to be inferred using either a
frequentist or a Bayesian approach.

@@

\includegraphics[width=0.8\textwidth,height=\textheight]{paper_files/figure-pdf/fig-rec1-1.pdf}

@@

\captionsetup{labelsep=none}

{\centering \includegraphics{mcmc_plot.jpg}

@@

\caption{: Posterior estimates of paramters for data simulated from the
proposed model with various combinations of
(\(\alpha\),\(\beta\),\(\varepsilon\)) and \(k_0=20\).}

\caption{\label{fig-rec2}}

@@

Turning now to real data, the goal is to fit the model to the degree
distributions of real networks from various sources. {[}describe
networks being used{]}. Alongside fitting this model to the degree
distributions, we then compare the fit to that of an mixture
distribution that was used in {[}clement{]} and is similar to others
that have been used to model degree distributions {[}others{]}.

@@

\includegraphics[width=0.8\textwidth,height=\textheight]{paper_files/figure-pdf/fig-real1-1.pdf}

@@

\includegraphics[width=0.8\textwidth,height=\textheight]{paper_files/figure-pdf/fig-shapes-1.pdf}

@@

\includegraphics[width=0.8\textwidth,height=\textheight]{paper_files/figure-pdf/fig-pa-1.pdf}

@@

\captionsetup{labelsep=none}

{\centering \includegraphics{pars_plot.png}

@@

\caption{: Posterior estimates of paramters for real data.}

\caption{\label{fig-rec2}}

@@

If this were the case, the posterior distributions of parameters from
Section~\ref{sec-real} could be interpreted as the approximate
distributions from which a vertex's fitness values are drawn instead of
providing estimates of the universal parameter values. Additionally, it
allows the ability to set some parameters as universal and some as
vertex-wise for example one could let \(\alpha, \beta, k_0\) be the same
for all vertices but let each vertex in the network have its own ``zero
appeal''
\(\varepsilon_i\overset{\text{iid}}{\sim} \text{Ga}(\theta_1, \theta_2)\);
moving the model to being more realistic and accounting for the fact
that in many real networks the objects vertices represent have some
value instrinsic to themselves other than the number of connections they
have.
